---
title: "Myc;Ptenfl TNBC mouse model RNA-seq analysis"
author: "Zinab O. Doha"
date: "01/28/2021"
output: html_document
---

# Import Libraries

```{r}
library(tidyverse)
```



# Import and format data

```{r}
data1 <- read.csv("expression_file.txt", sep = '\t' )

mouse_groups <- c(0, 0, 0, 0, 0, 0, 1, 1, 1, 0, 0, 1, 1) ## class labels
mouse_groups <- as.factor(mouse_groups)

rownames(data1) <- data1$genes ## row names should be gene names
head(data1) ## preview data
data1 <- data1 %>% select(-genes) ## remove gene names column -- now set as row labels
```


```{r}
mat1 <- data.matrix(data1) ## switch data type
t(head(mat1))
```

# Analysis  
## Plot samples by first two principal components

```{r}
pca1 <- prcomp(mat1) ## run PCA
pca2 <- as.data.frame(pca1$rotation) ## take PCS
pca3 <- pca2 %>% select(PC1, PC2) ## select first 2
pca3 <- cbind(pca3, mouse_groups) ## apply class labels
pca3 ## look at PCs
pca_plot <- ggplot(data = pca3, aes(x = PC1, y = PC2, colour = mouse_groups)) + geom_point(size = 1, alpha = .6, width = 0.2) ## create plot
pca_plot ## plot by first 2 principal components
```

## Normalize expression values and create heatmap from most variable loci

```{r}
full <- t(mat1) ## transpose matrix

max_vals <- c()

for (feat in seq(dim(full)[2])){ ## create vector of maximum expression values for every locus
  max_vals <- c(max_vals, max(full[, feat]))
}
```

## Normalize expression values

```{r}
samples <- c()

for (obs in seq(dim(full)[1])){ ## for every column, divide by maximum value of that feature to normalize between 0 and 1
  newrow <- full[obs,] / max_vals 
  samples <- rbind(samples, newrow)
}

samples[is.na(samples)] <- 0 ## replace NaNs with 0 -- fix divide by zero error

```

## Look for most variable sites

```{r}
variances <- c()

for (feat in seq(dim(samples)[2])){ ## calculate variance for every feature
  variances <- c(variances, var(full[, feat]))
}

```


```{r}
## take top "n" most variable loci

n = 1200 ## choose how many loci to include in heatmap

var_ordered <- variances[order(variances, decreasing = TRUE)] ## rank features by variance
cutoff <- min(var_ordered[1:n+1]) ## take "n" most variable sites


bool1 <- variances > cutoff ## boolean index for most variably expressed loci
```

## Select most variable sites 

```{r}
most_var <- samples[,bool1] ## index with most variable sites
rownames(most_var) <- rownames(full) ## sample names as row names
```

## Create heatmap / dendrograms 

```{r}
heatmap(most_var) ## create heatmap
```


SR: 1, 2, 3, 4, 5, 6, 10, 11

SP: 7, 8, 9, 12, 13

Developing dendrogram from 1200 most variably expressed genes, recapitulates 2 subgroups

# Human subtyping 

Parallels between four subtypes in human TNBC and mouse data


```{r}
#subtybing
#install.packages("reshape2")
#install.packages("ggplot2")
#install.packages("tidyr")
library(reshape2)
library(ggplot2)
library(tidyr)
```

 
## Read in files
```{r}
expression_file = "expression_file.txt"
df = read.csv(expression_file, sep="\t", header=TRUE, row.names=1, check.names=FALSE)
 
centroid_file = "Human_TNBC_77centroids.csv"
centroids <- read.csv(centroid_file, header=TRUE, row.names=1, sep="\t")
 
#names <- df[,1]
#rownames(df) = make.names(names, unique=TRUE)
#rownames(df) <- df[,1]
```

 
## Scale values and subset genes in signature
```{r}
df <- t(scale(t(df), center=TRUE, scale=TRUE))
```

 
## Make gene names uppercase
```{r}
names <- rownames(df)
rownames(df) = toupper(names)
```

 
## Replace NaNs with 0s
```{r}
df[is.na(df)] <- 0
 
df_data_filter <- df_data[df_data$Var1 %in% centroids$Gene, ]
 
centroids_filtered <- centroids[centroids$Gene %in% df_data_filter$Var1, ]
 
df2 <- df_data_filter %>%
  spread(Var2, Freq)
```

 
## x must be numeric, make first column numeric
```{r}
rownames(df2) <- df2$Var1
 
df2 <- subset(df2, select = -Var1)
```

 
## Divide df into sub data frames to make compatible dimensions
 
```{r}
df3 <- subset(df2, select = c(PT_Myc_PTEN_MGT_b1_TPM, PT_Myc_PTEN_MGT_b2_TPM, PT_Myc_PTEN_MGT_b3_TPM, PT_Myc_PTEN_MGT_b4_TPM))
df4 <- subset(df2, select = c(PT_Myc_PTEN_MGT_b5_TPM, PT_Myc_PTEN_MGT_b6_TPM, PT_Myc_PTEN_MGT_b7_TPM, PT_Myc_PTEN_MGT_b8_TPM))
df5 <- subset(df2, select = c(PT_Myc_PTEN_MGT_b9_TPM, PT_Myc_PTEN_MGT_b10_TPM, PT_Myc_PTEN_MGT_b11_TPM, PT_Myc_PTEN_MGT_b12_TPM))
df6 <- subset(df2, select = c(PT_Myc_PTEN_MGT_b11_TPM, PT_Myc_PTEN_MGT_b12_TPM, PT_Myc_PTEN_MGT_b13_TPM, PT_Myc_PTEN_MGT_b14_TPM))
```

 
## y must be numeric, make first column numeric
```{r}
rownames(centroids_filtered) <- centroids_filtered$Gene
centroids_filtered <- subset(centroids_filtered, select = -Gene)
```

 
## Run correletion
```{r}
cor <- cor(df3, centroids_filtered, method="pearson")
#cor <- cor(df4, centroids_filtered, method="pearson")
#cor <- cor(df5, centroids_filtered, method="pearson")
#cor <- cor(df6, centroids_filtered, method="pearson")

 
cor <- melt(as.matrix(cor))
colnames(cor) <- c("Sample", "Subtype", "Pearson")
```

 
## Bar plot of results

```{r}
ggplot(cor,aes(x=Sample, y=Pearson, fill=Subtype)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) +
  guides(fill=guide_legend(ncol=1))
 
ggsave(file = "subtype.cor4.png", device = "png", width = 5, height = 5, dpi=300)
```
## Save file
```{r}
saveRDS(cor,
        file = 'HUMAN_CORELATION.rds')
```

# Human corelation heatmap
```{r}
TNBC <- read.rds("HUMAN_CORELATION.rds", row.names = 1)


inputTN <- as.matrix(t(TNBC))



col_anno <- data.frame(histology = str_split(colnames(inputTN), pattern = '_', simplify = TRUE)[,1], row.names = colnames(inputTN))

histology        <- c("#FC717F", "#00A5FF")
names(histology) <- c("SP", "SR")
anno_colors <- list(histology = histology)
dcols = dist(t(inputTN), method = "minkowski")

pheatmap(prop.table(inputTN, margin = 2),scale="column", annotation = col_anno, annotation_colors = anno_colors, fontsize_row = 12, main = 'Human TNBC Spearman correlation (column scaled)', legend = FALSE)

pheatmap(inputTN, scale="column", annotation = col_anno, annotation_colors = anno_colors, fontsize_row = 16, main = 'Human TNBC Spearman correlation (column scaled)')

```


# frequency of subtypys (Pie shart)
```{r}
library(ggplot2)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(forcats)
library(scales)


pi <- read.csv("MycpSubtypes.csv")

bp<- ggplot(pi, aes(x="", y=value, fill=subtypes))+
geom_bar(width = 1, stat = "identity")
bp

pi1 <- ggplot(pi, aes(x="", y=value, fill=subtypes)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0)
pi1

pie <- bp + coord_polar("y", start=0) + geom_text(aes(label = label),
            position = position_stack(vjust = 0.5), size=10) +
  coord_polar(theta = "y") + theme(text = element_text(size = 4)) 
pie
pie + scale_fill_manual(values=c("#56B4E9", "#E69F00", "#CC6666", "#66CC99" )) + theme_void() +theme(text = element_text(size = 20)) 

```

# Cell-Stat
```{r}
stat <- read.csv("cell_stat_2.csv",row.names = 1)


inputS <- as.matrix(t(stat))

col_anno <- data.frame(histology = str_split(colnames(inputS), pattern = '_', simplify = TRUE)[,1], row.names = colnames(inputS))

histology        <- c("#FC717F", "#00A5FF")
names(histology) <- c("SP", "SR")
anno_colors <- list(histology = histology)

dcols_S = dist(t(inputS), method = "minkowski")

pheatmap(inputS, scale="row", annotation = col_anno, annotation_colors = anno_colors, fontsize_row = 12, color=colorRampPalette(c( "blue", "black", "RED"))(50),clustering_distance_cols = dcols_S,  cluster_rows = FALSE, cluster_cols = TRUE )


```

# GSEA
```{r}
gsea_sr <- read.csv("sr.csv")
```

## GSEA (shart)
```{r}
library(dplyr)
library(forcats)
gsea_sr %>%
  mutate(HALLMARK = fct_reorder(HALLMARK, NES)) %>%
  ggplot( aes(x=HALLMARK , y = NES , fill=Histology)) +
    geom_bar(stat="identity", position="identity")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 15))+
  coord_flip()
```

## GSEA (dotplot)
```{r}
dot_df = gsea_sr ## small dataset
dot_df$type = "upregulated"
dot_df$type[dot_df$NES < 0] = "downregulated"

## from Tommy's code
p <- ggplot(dot_df, aes(x = NES, y = HALLMARK)) + 
               geom_point(aes(size = SIZE, color = NOM_p_val)) +
               theme_bw(base_size = 14) +
        scale_colour_gradient(limits=c(0, 0.10), low="red") +
        ylab(NULL) +
        ggtitle("GO pathway enrichment")

p + facet_grid(.~type)

```

#mIHC heatmap 
```{r}
library("gplots")
library("RColorBrewer") 
library("pheatmap")

core_h <- read.csv("core.csv",row.names = 1)


inputc <- as.matrix(t(core_h))

col_anno <- data.frame(histology = str_split(colnames(inputc), pattern = '_', simplify = TRUE)[,1], row.names = colnames(inputc))


pheatmap(inputc,scale="row", annotation_col = col_anno, main = "pheatmap row annotation")
```

##Row scaling Core
```{r}
histology        <- c("#FC717F", "#00A5FF")
names(histology) <- c("SP", "SR")
anno_colors <- list(histology = histology)
pheatmap(inputc, scale="row", annotation = col_anno, annotation_colors = anno_colors, fontsize_row = 12, main = 'Core_row scaled')
```

## Row scaling_Border

```{r}
border <- read.csv("border.csv",row.names = 1)


inputb <- as.matrix(t(border))

col_anno <- data.frame(histology = str_split(colnames(inputb), pattern = '_', simplify = TRUE)[,1], row.names = colnames(inputb))

histology        <- c("#FC717F", "#00A5FF")
names(histology) <- c("SP", "SR")
anno_colors <- list(histology = histology)
dcols = dist(t(inputb), method = "minkowski")

pheatmap(inputb, scale="row", annotation = col_anno, annotation_colors = anno_colors, fontsize_row = 12, clustering_distance_cols = dcols, main = 'Border_row scaled' , legend = TRUE)
```

## Row scaling preiphry
```{r}
perph <- read.csv("perph.csv",row.names = 1)


inputp <- as.matrix(t(perph))

col_anno <- data.frame(histology = str_split(colnames(inputp), pattern = '_', simplify = TRUE)[,1], row.names = colnames(inputp))

histology        <- c("#FC717F", "#00A5FF")
names(histology) <- c("SP", "SR")
anno_colors <- list(histology = histology)
pheatmap(inputp, scale="row", annotation = col_anno, annotation_colors = anno_colors, fontsize_row = 12, main = 'Periphery_row scaled' )
```

## cytokines heatmap
```{r}
cyt <- read.csv("cytokines.csv",row.names = 1)


inputcy <- as.matrix(t(cyt))

col_anno <- data.frame(histology = str_split(colnames(inputcy), pattern = '_', simplify = TRUE)[,1], row.names = colnames(inputcy))

histology        <- c("#FC717F", "#00A5FF")
names(histology) <- c("SP", "SR")
anno_colors <- list(histology = histology)

dcols_cy = dist(t(inputcy), method = "minkowski")

pheatmap(inputcy, scale="row", annotation = col_anno, annotation_colors = anno_colors, fontsize_row = 12, clustering_distance_cols = dcols_cy, color=colorRampPalette(c("navy", "white", "yellow"))(50) )
```

# Metastisis Rate
```{r}
library(ggplot2)
meta <- read.csv("meta.csv")
my_ggp <- ggplot (data = meta,
       aes (x = Tumor,
            y = Metastisis_rate, fill = Tumor)) +
           geom_bar(stat = "Identity") +
  ggtitle("Metastisis Rate")
my_ggp 

my_ggp + theme(text = element_text(size = 18
                                  )) + geom_violin(colour = "yellow", alpha = 0.1)
```

```{r}